Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-27 draft.tex
Revised Version from Steffen, new Style

Cohort studies comprise of heterogenous variables describing the medical condition as well as demographic and lifestyle factors and, more recently, medical image data.

@@

Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of people, randomly selected to avoid any bias.
%
The subjects are selected without focus on a certain disease.
%
A large cohort size is essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
% TODO - Khonya Paper Venue
% TODO - Introduction Highlight Overview visualization?
% TODO - Klemm et al. - Klingt ein bisschen shizophren
Cohort studies often include medical image data.
% 
The concurrent analysis of image data and non-spatial epidemiological factors requires techniques that reach beyond standard statistical methods.
%
For instance, segmentation of the image data is required for an analysis of anatomical structure \com{and of possible} correlations between this structure and epidemiological factors.
%
Semi-automatic segmentation techniques are promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
% Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
% %
% Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%

@@

Angelelli et al. \cite{Angelelli2014} focus on a data organization for the interactive visual analysis of heterogeneous cohort study data.
%
%\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
%
Brain image data was integrated into the analysis by first segmenting brain regions and tracking neural pathways and then deriving attributes from both, e.g., volume and fractional anisotropy.
%
The data-cube model facilitates the seamless integration of heterogeneous image-based and non-image data.
%
A multiple coordinated view framework linked spatial and non-spatial data views.
%
%The approach is a demonstrated by a case-study analysis of selected aspects of brain connectivity.
% --- Steffens Fazit ---
%
Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
%
While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple views showing the information of subgroups and their respective deviation from the entire cohort.
%
%\rem{We also integrate the data-driven definition of subgroups using clustering, statistically-based suggestions of potentially interesting features, and adaptive feature visualization.} %tailored to epidemiological data, e.g., a mosaic plot for the frequent categorical variables.
% %
% %Another unique characteristic is our web-based implementation and the integration of D3.js.
% %
% %The latter offers an extensive set of information visualization plots, which are at the disposal of our system.
% %
% --- / Steffens Fazit ---

@@

Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or whether they employ statistical measures.

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image-Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
%Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
%Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on variables associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
%
Variables not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive measures, such as diameters.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
%It is easily possible that omitted information would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and could also identify possible anatomical confounders--an epidemiolgical research result in itself.
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative variable selection process (see Fig.~\ref{fig:WorkflowComparison}~b).
%\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

@@

% \subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%
The suitability of visualization techniques for epidemiological data depends on its ability to compare multiple data variables while highlighting associations.
%
%The methods have to reflect the routines that epidemiologists take in order to be incorporated into their research.
%
Visual evaluations of data are therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of the system.
% \subsection{System Structure and Design} \label{Structure and Workflow}
\subsection{Design and Visualization Techniques} \label{Structure and Workflow}
% \begin{itemize}
% 	\item [x] Was sind die requirements des tools?
% 	\item [x] how it was decided to create this particular design
% 	\item [x] which kind of User interaction was made?
% 	\item [x] Spezielle Anforderungen durch Webseitenformat, weggehen von allgemeinen UI Elementen wie Menubar, keine Rechtsklickmenüs.
% 	\item [x] Entschieden eine saubere UI zu bauen, Unterscheide durch Whitespace akzentuieren. Interface in Fig. 3 zu sehen.
% 	\item [x] Wie wird interagiert, wie wird gebrusht, was wird gelinkt?
% 	\item [x] wie wird das Clustering initialisiert?
% 	\item auf Kontingenzmatrix mit verweisen als Teil des Interfaces (dieses auch noch anpassen)
% \end{itemize}
%Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
Early it became clear that we have to rely a lot on online communication due to the large spatial distance towards each other.
%
Hence, we build our system using web technologies. %, described in detail in Section~\ref{implementation}.
%
By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts in the development process with little effort.
%
Incorporating the \emph{IVA} workflow for image-centric cohort study data requires \emph{overview visualizations} as well as \emph{multivariate visualizations}, which bring image-derived information in context to non-image variables.

The focus on web technologies is not without tradeoffs.
%
Classical UI elements, such as the menu bar or custom right-click menus, are technically possible, but not common in this domain.
%
In favor of a clean layout, we designed the system without such components.
%
Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we tried to make the interface as minimalistic as possible, treating the resulting space as \emph{canvas} for the data.
%
% \subsubsection{System Structure}
We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize} \itemsep0.2em
		\item The \emph{sidebar}, which contains all epidemiological variables. The cluster results group variables like categorical variables and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
	%
	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (Fig.~\ref{fig:visualization}~b).
	%
	\item The interactive \emph{pivot table} gives detailed numerical information of the variables in the canvas view. This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization}~c).
	%
	\item The \emph{contingency view} depicts relations for variables in the canvas in an contingency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
\end{itemize}

@@

\paragraph{Adaptive Variable Visualization.} \label{sec:AdaptiveFeatureVisualization}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Contingency matrix of 129 variables (127 data set variables, 2 cluster results) showing 16,641 combinations.
 %
 Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
 %
 Color brightness encodes association strength.
 %Mouse-over focuses a entry and highlights the feature names for better readability.
 Moving the mouse over an entry enlarges the variable names for better readability.
 % Potential Cut here
 The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}
The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the variable types and number to allow for \emph{multivariate analysis}.
%
Categorical data are either mapped to bar charts (single variables) or mosaic plots (multiple variables).
%
Figure \ref{fig:visualization} describes this dynamic adjustment.
%
Continuous data can be visualized using scatter plots (two variables) or parallel coordinates (multiple variables), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
%
Since the number of categories often depends on the hypothesis, the discretization steps can be adapted dynamically.
%
Too many groups potentially generate sparse bins not suited for statistical evaluation.
%
Not enough groups overgeneralize information.
%
Adaptive discretization is an option, but imposes possible overfitting to the data.
%
Conclusions based on statistical relationships derived from groups already biased by variable distribution are heavily influenced by the used discretization.
%
Therefore, we follow the convention to use bins of equal size.
%

Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, information derived from the medical image data are incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization}~b).
%
The 3D plots can be navigated using standard mouse input, the camera is synchronized between all views to enable direct comparison.
%
The distance from a group mean shape is mapped to the global mean using color.
%
This allows to assess local shape changes (Fig.~\ref{fig:visualization}) and is an important information to the epidemiologist.
%
Until now, epidemiologists were not able to rapidly validate shape differences based on non-image variables.
%
Dropping a variable on an existing plot adapts the visualization dynamically to allow for comparison (Fig.~\ref{fig:visualization}~right).
% --- PARAGRAPH ---

@@

\paragraph{Automated Variable Suggestion using a Contingency Matrix.}
%\rem{As previously discussed}
Highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern for analyzing variables outside the shape space.
%
Turkay et al. \cite{Turkay2013} used the approach to calculate key figures based on the distribution functions of each variable derived from the image data.
%
Since the majority of our data are categorical variables, we have to employ different solutions.
%
The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
%
It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
%
\emph{Cram\'{e}r's V} is defined as:
\begin{equation}
V = \sqrt{\frac{X^2}{N(k-1)}},
\end{equation}
where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ is either the row or column count, depending on which one is lower.
%
$V$ yields values between $0$, meaning that two variables are completely independent, and $1$ indicates that they are the same.
%
%Since \emph{Cram\'{e}r's V} is always positive, 
\emph{Cram\'{e}r's V} does not allow to infer the dependency direction.
%

It shares the same restrictions as \emph{Pearson's $X^2$}.
%
The expected counts in the contingency table have to be larger than 5 for $80\%$ of the entries and no expected value must be smaller than one \cite{Cochran1952}.
%
Some manifestations and variable combinations, which are only exposed by small subject groups, cannot be assessed with this technique.
%
They cannot be included into the epidemiological analysis, since statistical validation needs a minimum count to be valid.
%
The contingency matrix highlights correlations between all variables. %of visualized features with features in the data set.
%
This aims to highlight variables possibly associated with the focused hypothesis and to trigger new hypotheses.
%
Contingency is visualized using an interactive contingency matrix with association power mapped to color brightness.
%
The distinction whether an association is a confounder or an effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
%
The contingency matrix visualization is an \emph{overview visualization}--something the epidemiological community lacks and is in great need of.

@@

Subject clustering is triggered automatically as \emph{local investigation} for each variable after it was added to the canvas by the user.

@@

	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance result in many open source libraries, which are well documented and driven by active communities. We use \emph{WebGL} for rendering shape information.

@@

%These advantages do not come without drawbacks.

@@

However, sophisticated libraries/languages, such as the \texttt{Visualization Toolkit}\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}} for statistics, are either not available at all or only accessible through complex client-server systems.

@@

\subsubsection{Data Preprocessing} \label{application:Data Preprocessing}
The data are preprocessed as described in Section~\ref{Data Preprocessing}.
%
\paragraph{Non-Image Data.}

To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the \texttt{JSON} file format.
%
Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the \texttt{SAS v9+} format, which saves the data labels, and exported the data values as non-labeled \texttt{CSV}.
%
A short script combined both data sources to a \texttt{JSON} file.
%
The data types had to be transferred manually.
%
Each variable is stored as an object containing the data as array, its data type, a detailed description and data dictionary translating value or error IDs to values.
% \begin{itemize}
% 	\item the data as array of values--categorical data and error codes are stored using IDs,
% 	\item the data type (continuous, nominal, ordinal, dichotomous),
% 	\item a detailed description of the feature, and
% 	\item the data dictionary translating value or error IDs to values.
% \end{itemize}
% Each feature is stored as an object containing:
% \begin{itemize}
% 	\item the data as array of values--categorical data and error codes are stored using IDs,
% 	\item the data type (continuous, nominal, ordinal, dichotomous),
% 	\item a detailed description of the feature, and
% 	\item the data dictionary translating value or error IDs to values.
% \end{itemize}
%
% Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
%
According to epidemiological publications, we set the number of groups to five, to allow for contingency assessment.

\paragraph{Image Data.} \label{Image-Data}
The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
%
Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
%
It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
%
The registered models capture resilient information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
%
Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
%
We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
%
For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).

>>>>

Cohort studies comprise heterogenous variables describing the medical condition as well as demographic and lifestyle factors and, more recently, medical image data.

@@

Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of people, randomly selected to avoid any bias.
%
The subjects are selected without focus on a certain disease.
%
A large cohort size is essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
% TODO - Khonya Paper Venue
% TODO - Introduction Highlight Overview visualization?
% TODO - Klemm et al. - Klingt ein bisschen shizophren
Cohort studies often include medical image data.
% 
The concurrent analysis of image data and non-spatial epidemiological factors requires techniques that reach beyond standard statistical methods.
%
For instance, segmentation of the image data is required for an analysis of anatomical structure and of possible correlations between this structure and epidemiological factors.
%
Semi-automatic segmentation techniques are promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
% Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
% %
% Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%

@@

Angelelli et al. \cite{Angelelli2014} focus on the data organization for an interactive visual analysis of heterogeneous cohort study data.
%
The proposed data-cube model facilitates the seamless integration of image-based and non-image data.
%\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
%
In a demonstration of the model, brain image data was integrated into the analysis by first segmenting brain regions and tracking neural pathways and then deriving attributes from both, e.g., volume and fractional anisotropy.
%
A multiple coordinated view framework then linked spatial and non-spatial data views.
%
%The approach is a demonstrated by a case-study analysis of selected aspects of brain connectivity.
% --- Steffens Fazit ---
%
Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
%
While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple views showing the information of subgroups and their respective deviation from the entire cohort.
%
%\rem{We also integrate the data-driven definition of subgroups using clustering, statistically-based suggestions of potentially interesting features, and adaptive feature visualization.} %tailored to epidemiological data, e.g., a mosaic plot for the frequent categorical variables.
% %
% %Another unique characteristic is our web-based implementation and the integration of D3.js.
% %
% %The latter offers an extensive set of information visualization plots, which are at the disposal of our system.
% %
% --- / Steffens Fazit ---

@@

Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects' works or whether they employ statistical measures.

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image-Centric Cohort Study Data in an Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
%Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
%Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on variables associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
%
Variables not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive measures, such as diameters.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
%It is easily possible that omitted information would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and could also identify possible anatomical confounders--an epidemiolgical research result in itself.
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative variable selection process (see Fig.~\ref{fig:WorkflowComparison}~b).
%\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

@@

% \subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%
The suitability of visualization techniques for epidemiological data depends on their ability to compare multiple data variables while highlighting associations.
%
%The methods have to reflect the routines that epidemiologists take in order to be incorporated into their research.
%
Visual evaluations of data are therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of our system.
% \subsection{System Structure and Design} \label{Structure and Workflow}
\subsection{Design and Visualization Techniques} \label{Structure and Workflow}
% \begin{itemize}
% 	\item [x] Was sind die requirements des tools?
% 	\item [x] how it was decided to create this particular design
% 	\item [x] which kind of User interaction was made?
% 	\item [x] Spezielle Anforderungen durch Webseitenformat, weggehen von allgemeinen UI Elementen wie Menubar, keine Rechtsklickmenüs.
% 	\item [x] Entschieden eine saubere UI zu bauen, Unterscheide durch Whitespace akzentuieren. Interface in Fig. 3 zu sehen.
% 	\item [x] Wie wird interagiert, wie wird gebrusht, was wird gelinkt?
% 	\item [x] wie wird das Clustering initialisiert?
% 	\item auf Kontingenzmatrix mit verweisen als Teil des Interfaces (dieses auch noch anpassen)
% \end{itemize}
%Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
Early it became clear that we have to rely a lot on online communication due to the large spatial distance towards each other.
%
Hence, we built our system using web technologies. %, described in detail in Section~\ref{implementation}.
%
By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts in the development process with little effort.
%
Incorporating the \emph{IVA} workflow for image-centric cohort study data requires \emph{overview visualizations} as well as \emph{multivariate visualizations}, which bring image-derived information in context to non-image variables.

The focus on web technologies is not without tradeoffs.
%
Classical UI elements, such as the menu bar or custom right-click menus, are technically possible, but not common in this domain.
%
In favor of a clean layout, we designed the system without such components.
%
Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we designed the interface as minimalistic as possible, treating the resulting space as \emph{canvas} for the data.
%
% \subsubsection{System Structure}
We divided the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize} \itemsep0.2em
		\item The \emph{sidebar}, which contains all epidemiological variables. The cluster results group variables like categorical variables and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
	%
	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (Fig.~\ref{fig:visualization}~b).
	%
	\item The interactive \emph{pivot table} gives detailed numerical information of the variables in the canvas view. This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization}~c).
	%
	\item The \emph{contingency view} depicts relations for variables in the canvas in an contingency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
\end{itemize}

@@

\paragraph{Adaptive Variable Visualization.} \label{sec:AdaptiveFeatureVisualization}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Contingency matrix of 129 variables (127 data set variables, 2 cluster results) showing 16,641 combinations.
 %
 Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
 %
 Color brightness encodes association strength.
 %Mouse-over focuses a entry and highlights the feature names for better readability.
 Moving the mouse over an entry enlarges the variable names for better readability.
 % Potential Cut here
 The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 %
 The contingency pane is not shown here.
 }
 \label{fig:similarity}
\end{figure}
The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the variable types and number to allow for \emph{multivariate analysis}.
%
Categorical data are either mapped to bar charts (single variables) or mosaic plots (multiple variables).
%
Figure \ref{fig:visualization} describes this dynamic adjustment.
%
Continuous data can be visualized using scatter plots (two variables) or parallel coordinates (multiple variables), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
%
Since the number of categories often depends on the hypothesis, the discretization steps can be adapted dynamically.
%
Too many groups potentially generate sparse bins not suited for statistical evaluation.
%
Not enough groups overgeneralize information.
%
Adaptive discretization is an option, but imposes possible overfitting to the data.
%
Conclusions based on statistical relationships derived from groups already biased by variable distribution are heavily influenced by the used discretization.
%
Therefore, we follow the convention to use bins of equal size.
%

Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, information derived from the medical image data are incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization}~b).
%
The 3D plots can be navigated using standard mouse input, the camera is synchronized between all views to enable direct comparison.
%
The distance from a group mean shape is mapped to the global mean using color.
%
This allows to assess local shape changes (Fig.~\ref{fig:visualization}) and is an important information to the epidemiologist.
%
Until now, epidemiologists were not able to inspect shape differences based on non-image variables.
%
Dropping a variable on an existing plot adapts the visualization dynamically to allow for comparison (Fig.~\ref{fig:visualization}~right).
% --- PARAGRAPH ---

@@

\paragraph{Automated Variable Suggestion using a Contingency Matrix.}
%\rem{As previously discussed}
Highlighting potentially interesting associations in the data set is one major benefits of the \emph{IVA}-powered approach and is part of the \emph{multivariate analysis} pattern for analyzing variables outside the shape space.
%
Turkay et al. \cite{Turkay2013} used the approach to calculate statistical key figures based on the distribution functions of each variable derived from the image data.
%
Since the majority of our data are categorical variables, we have to employ different solutions.
%
The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
%
It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
%
\emph{Cram\'{e}r's V} is defined as:
\begin{equation}
V = \sqrt{\frac{X^2}{N(k-1)}},
\end{equation}
where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ is either the row or column count, depending on which one is lower.
%
$V$ yields values between $0$, meaning that two variables are completely independent, and $1$ indicates that they are the same.
%
%Since \emph{Cram\'{e}r's V} is always positive, 
\emph{Cram\'{e}r's V} does not allow to infer the dependency direction.
%

It shares the same restrictions as \emph{Pearson's $X^2$}.
%
The expected counts in the contingency table have to be larger than five for $80\%$ of the entries and no expected value must be smaller than one \cite{Cochran1952}.
%
Some manifestations and variable combinations, which are only exposed by small subject groups, cannot be assessed with this technique.
%
They cannot be included into the epidemiological analysis, since statistical validation needs a minimum count to be valid.
%
The contingency matrix highlights correlations between all variables. %of visualized features with features in the data set.
%
This aims to highlight variables possibly associated with the focused hypothesis and to trigger new hypotheses.
%
Contingency is visualized using an interactive contingency matrix with association power mapped to color brightness.
%
The distinction whether an association is a confounder or an effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
%
The contingency matrix visualization is an \emph{overview visualization}--something the epidemiological community lacks and is in great need of.

@@

Subject clustering is triggered automatically as \emph{local investigation} for a variable after it was added to the canvas by the user.

@@

	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance results in many open source libraries, which are well documented and driven by active communities. We use \emph{WebGL} for rendering shape information.

@@

These advantages do not come without drawbacks.

@@

Sophisticated libraries/languages, such as the \texttt{Visualization Toolkit}\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}} for statistics, are either not available at all or only accessible through complex client-server systems.

@@

\subsubsection{Data Preprocessing} \label{application:Data Preprocessing}
The data processing follows the description in Section~\ref{Data Preprocessing}.
%
\paragraph{Non-Image Data.}

To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the \texttt{JSON} file format.
%
Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the \texttt{SAS v9+} format, which saves the data labels, and exported the data values as non-labeled \texttt{CSV}.
%
A short script combined both data sources to a \texttt{JSON} file.
%
The data types had to be transferred manually.
%
Each variable is stored as an object containing the data as array, its data type, a detailed description and data dictionary translating value or error IDs to values.
% \begin{itemize}
% 	\item the data as array of values--categorical data and error codes are stored using IDs,
% 	\item the data type (continuous, nominal, ordinal, dichotomous),
% 	\item a detailed description of the feature, and
% 	\item the data dictionary translating value or error IDs to values.
% \end{itemize}
% Each feature is stored as an object containing:
% \begin{itemize}
% 	\item the data as array of values--categorical data and error codes are stored using IDs,
% 	\item the data type (continuous, nominal, ordinal, dichotomous),
% 	\item a detailed description of the feature, and
% 	\item the data dictionary translating value or error IDs to values.
% \end{itemize}
%
% Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
%
Following epidemiological publications, we set the number of groups to five, to allow for contingency assessment.

\paragraph{Image Data.} \label{Image-Data}
The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
%
Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
%
The model uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
%
Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
%
Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
%
We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
%
For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature)  \cite{Klemm2013VMV}.
